Intrusion Detection with Genetic Algorithms and Fuzzy Logic

### 2013-11-12 paper.tex
Made some changes based on feedback.

\section{Introduction}
An attack on important and confidential data is a concern for many people, so it is important to have a way to detect and analyze these attacks. A way of detecting attacks is by using an intrusion detection system. An intrusion detection system is a device that monitors network activities for malicious or abnormal behaviors and then produces reports~\cite{DBLP:journals/corr/abs-1204-1336}. There are various ways of training the intrusion detection system about the possible threats. Two different ways, which I will talk about in this paper, are the use of genetic algorithms and fuzzy logic.

In Section~\ref{sec:background} I will give some background on intrusion detection. I will explain the main types of networking attacks in Section~\ref{sec:typesAttacks}, and ways of detecting attacks in Section~\ref{sec:detectionMeth}. In Section~\ref{sec:dataSets} I will explain two data sets that are used in intrusion detection. I will explain what \emph{rules} are in Section~\ref{sec:rules}, and then give information on fuzzy logic and genetic algorithms in Sections~\ref{sec:fuzzyLogic} and~\ref{sec:GA}. After that, I will explain the measures that are used in determining the accuracy of an intrusion detection system in Section~\ref{sec:PosNeg}. In Section~\ref{sec:genAlgImp} I will explain the use of a genetic algorithm in an intrusion detection system, and in Section~\ref{sec:fuzGenAlgImp} I will explain the use of a fuzzy genetic algorithm. Section~\ref{sec:conclusion} has some conclusions about the use of genetic algorithms and fuzzy genetic algorithms in intrusion detection systems.

@@

Denial of service (DoS) attacks happen when an attacker makes a machine inaccessible to a user by making it too busy to serve legitimate requests. For example, many systems lock out a user from an account after a certain number of failed login attempts. An attacker would be able to use this to prevent legitimate users from logging in.~\cite{DoSAttacks} Remote to user (R2L) attacks happen when an attacker sends packets to a machine over the network in order to gain access to things a local user would have on the machine. An example is when an attacker tries to gain access to a machine by guessing possible usernames and passwords. User to root (U2R) attacks happen when an attacker starts out with access on the machine and then tries to gain root access to the system. For example, if a program expects a user to input their name, the programmer has to decide how many characters that name buffer will require. Assume the program allocates 20 characters for the name buffer. Suppose the user's name has 35 characters. The last 15 characters will overflow the buffer, which will then overwrite the instructions that are to be executed next. An attacker can cause commands to be executed by manipulating the data that overflows. Probing happens when an attacker examines a machine in order to collect information about weaknesses or vulnerabilities that in the future could be used to compromise the system.~\cite{DBLP:journals/corr/abs-1204-1336, typesOfAttacks}

@@

An anomaly is something that deviates from what is normal. Anomaly-based detection looks for patterns of activity that are rare and uncommon. It is an effective way to detect new attacks. Anomaly-based detection is also called \emph{behavior-based detection}.~\cite{DBLP:journals/corr/abs-1204-1336}

\subsection{Data Sets}
\label{sec:dataSets}
Two different data sets are used in this paper: KDD99 and RLD09. The KDD99 data set is a benchmark data set that was generated by simulating a military network environment in 1999~\cite{6559603}. It has long been a standard data set for intrusion detection. The data in the set is classified as normal or attack activity. KDD99 uses 41 \emph{features}, which are properties of a \emph{record} (either an attack or normal activity), that are used to describe the activity.

The authors of~\cite{6496342, 6559603} used 8 out of the 41 KDD99 features~\cite{KDD99} in their system: 
\begin{enumerate}
  \item duration: the length of the normal or attack activity in seconds.
  \item src\_bytes: the number of bytes sent from source to destination. Source is the user who may or may not be an attacker, and destination is the server being potentially attacked.
  \item num\_failed\_logins: the number of failed login attempts.
  \item root\_shell: returns 1 if root shell is obtained, which means that the user is able to log in as root. This gives them the ability to do things like add accounts and change user passwords. If root shell is not obtained, 0 is returned.
  \item num\_access\_files: the number of operations on access control files. Access control files specify which users are granted access to objects and what operations are allowed on the objects. An example would be (Alice, delete), which would give Alice permission to delete the file.~\cite{accessControl}
  \item srv\_count: the number of connections to the same service as the current connection in the past two seconds.
  \item serror\_rate: the percentage of connections that have ``SYN" errors. When a client attempts to connect to a server, it first sends a SYN (synchronize) message to the server. The server then acknowledges the request by sending a SYN-ACK to the client. The connection is established when the client sends an ACK back to the server. A SYN error is a failure that happens early in this process.~\cite{TCP}
  \item same\_srv\_rate: the percentage of connections to the same service.
\end{enumerate}

The KDD99 data set is 14 years old, and newer attack types are not included in it because of its age. Because of this, the authors of~\cite{6496342, 6559603} created their own data set, RLD09, to use in their experiments. To create the data set, the authors captured network data from the Computer Engineering Department at King Mongkut's University of Technology Thonburi, in Bangkok, Thailand. The data has around ten million preprocessed data \emph{packets}. A packet is a unit of data that is carried throughout a network~\cite{networkPacket}. RLD09 has 17 different types of attacks that can be divided into denial of service attacks and probe attacks. It also has normal network activity. RLD09 uses 12 features, which include the number of packets, source ports, and destination ports.

@@

\subsection{Fuzzy Logic}
\label{sec:fuzzyLogic}
Attacks on systems do not always have a fixed pattern, so fuzzy logic is used to detect patterns that have a behavior that is between normal and unusual. A fuzzy logic rule has the following format: If <\emph{condition}> then <\emph{consequence}>, where \emph{condition} is a fuzzy variable, and \emph{consequence} is a fuzzy set. An example of this is: if the number of packets with the same destination address is high, then the pattern is unusual. To determine what is considered high, the values of the packets are divided into fuzzy sets. In Figure~\ref{fig:triangleFigure} there are three sets: low (region A), medium (region B), and high (region C). 
\begin{figure}
\centering
\psfig{file=triangleFigure.JPG}
\caption{Fuzzy shape with three sets: low (A), medium (B), and high (C)~\cite{DBLP:journals/corr/abs-1204-6416}}
\label{fig:triangleFigure}
\end{figure}
The x-axis are the values in the fuzzy set (number of packets), and the y-axis is the membership function, which determines the region. For example, if the number of packets with the same destination address is 15 then this will be considered as low for a degree of 0.4, but will be considered medium or high for other degrees. If the number of packets is determined to be high, then the connection will be aborted.~\cite{DBLP:journals/corr/abs-1204-6416}

\vspace{0.30cm}

@@

\subsection{Algorithm Overview}
The authors of~\cite{DBLP:journals/corr/abs-1204-1336} used a genetic algorithm to make their intrusion detection system. The system is divided into two phases: a precalculation phase and a detection phase. In the precalculation phase, a set of chromosomes are created using training data. Then this set of chromosomes is used in the detection phase for comparison. Algorithm~\ref{alg:precalc} is used in the precalculation phase.

@@

    chromosome}
  \ENDFOR
  \STATE{Remove some chromosomes with worse fitness}
  \STATE{Apply crossover to the selected pair of chromosomes of the population}
  \STATE{Apply mutation to each chromosome of the population}
\ENDWHILE
\end{algorithmic}
\end{algorithm}

@@

  \STATE{compare the predicted result with actual result}
  \STATE{find $A$, $B$, $\alpha$, and $\beta$}
  \STATE{// $A$ is total number of attack records. $B$ is total number of normal records. $\alpha$ is total number of attack records correctly identified as attack. $\beta$ is total number of normal records incorrectly classified as attack.}
\ENDFOR
\STATE{calculate fitness}
\STATE{//create next generation}
\STATE{preserve\_best()}
\STATE{crossover()}
\STATE{mutation()}
\end{algorithmic}
\end{algorithm}

The fitness function to be maximized is:
\begin{equation*}
\textrm{fitness function} = \frac{\alpha}{A} - \frac{\beta}{B}
\end{equation*}
In this implementation, a population size of 10 was used for each generation. An individual in the population represents a possible detection rule. The two best individuals from a present generation are preserved for the next generation. The other individuals in the new generation come from mutation and single-point crossover.

@@

The authors of~\cite{6496342, 6559603} first trained and tested the fuzzy genetic algorithm with the KDD99 data set. There were 6 different types of denial of service attacks and 4 different types of probe attacks. The detection rate of the KDD99 data set was 98.72\%. Then 26,500 records of the RLD09 data set were used as the training set. The detection rate was 97.97\%. The results of this experiment are shown in Table~\ref{tab:bothSetsResults}.

>>>>

\section{Introduction}
An attack on important and confidential data is a concern for many people, so it is important to have a way to detect and analyze these attacks. A way of detecting attacks is by using an intrusion detection system. An intrusion detection system is a device that monitors network activities for malicious or abnormal behaviors and then produces reports and alerts~\cite{DBLP:journals/corr/abs-1204-1336}. There are various ways of training the intrusion detection system about possible threats. Two approaches that I will talk about in this paper are the use of genetic algorithms and fuzzy logic.

In Section~\ref{sec:background} I will give some background on intrusion detection. I will explain the main types of networking attacks in Section~\ref{sec:typesAttacks}, and ways of detecting attacks in Section~\ref{sec:detectionMeth}. In Section~\ref{sec:dataSets} I will describe two data sets that are used in intrusion detection. I will explain what \emph{rules} are in Section~\ref{sec:rules}, and then give information on fuzzy logic and genetic algorithms in Sections~\ref{sec:fuzzyLogic} and~\ref{sec:GA}. After that, I will explain the measures that are used in determining the accuracy of an intrusion detection system in Section~\ref{sec:PosNeg}. In Section~\ref{sec:genAlgImp} I will explain the use of a genetic algorithm in an intrusion detection system, and in Section~\ref{sec:fuzGenAlgImp} I will explain the use of a fuzzy genetic algorithm. Section~\ref{sec:conclusion} has some conclusions about the use of genetic algorithms and fuzzy genetic algorithms in intrusion detection systems.

@@

Denial of service (DoS) attacks happen when an attacker makes a machine inaccessible to a user by making it too busy to serve legitimate requests. For example, many systems lock out a user from an account after a certain number of failed login attempts. An attacker would be able to use this to prevent legitimate users from logging in~\cite{DoSAttacks}. Remote to user (R2L) attacks happen when an attacker sends packets to a machine over the network in order to gain access to things a local user would have on the machine. An example is when an attacker tries to gain access to a machine by guessing possible usernames and passwords. User to root (U2R) attacks happen when an attacker starts out with access on the machine and then tries to gain root access to the system. For example, if a program expects a user to input their name, the programmer has to decide how many characters that name buffer will require. Assume the program allocates 20 characters for the name buffer. Suppose the user's name has 35 characters. The last 15 characters will overflow the buffer, which could then overwrite the instructions that are to be executed next. An attacker can cause commands to be executed by manipulating the data that overflows. Probing happens when an attacker examines a machine in order to collect information about weaknesses or vulnerabilities that in the future could be used to compromise the system.~\cite{DBLP:journals/corr/abs-1204-1336, typesOfAttacks}

@@

Anomaly-based detection looks for patterns of activity that are rare and uncommon. It is an effective way to detect new attacks. Anomaly-based detection is also called \emph{behavior-based detection}.~\cite{DBLP:journals/corr/abs-1204-1336}

\subsection{Data Sets}
\label{sec:dataSets}
Two different data sets are used in this paper: KDD99 and RLD09. The KDD99 data set is a benchmark data set that was generated by simulating a military network environment in 1999~\cite{6559603}. It has long been a standard data set for intrusion detection. The data in the set is classified as normal or attack activity. KDD99 uses 41 \emph{features}, which are properties of a \emph{record} (either an attack or normal activity), that are used to describe the activity and help to distinguish normal connections from attacks.

In research conducted by P. Jongsuebsuk, N. Wattanapongsakorn, and
C. Charnsripinyo~\cite{6496342, 6559603}, 8 out of the 41 KDD99 features~\cite{KDD99} were used in their system:
\begin{enumerate}
  \item duration: the length of the normal or attack activity in seconds.
  \item src\_bytes: the number of bytes sent from source to destination. Source is the user who may or may not be an attacker, and destination is the server being potentially attacked.
  \item num\_failed\_logins: the number of failed login attempts.
  \item root\_shell: returns 1 if root shell is obtained, which means that the user is able to log in as root. This gives them the ability to do things like add accounts and change user passwords. If root shell is not obtained, 0 is returned.
  \item num\_access\_files: the number of operations on access control files. Access control files specify which users are granted access to objects and what operations are allowed on the objects. An example would be (Alice, delete), which would give Alice permission to delete the file.~\cite{accessControl}
  \item srv\_count: the number of connections to the same service as the current connection in the past two seconds.
  \item serror\_rate: the percentage of connections that have ``SYN" errors. When a client attempts to connect to a server, it first sends a SYN (synchronize) message to the server. The server then acknowledges the request by sending a SYN-ACK to the client. The connection is established when the client sends an ACK back to the server. A SYN error is a failure that happens early in this process.~\cite{TCP}
  \item same\_srv\_rate: the percentage of connections to the same service.
\end{enumerate}

The KDD99 data set is 14 years old, and newer attack types are not included in it because of its age. Because of this, P. Jongsuebsuk, N. Wattanapongsakorn, and
C. Charnsripinyo~\cite{6496342, 6559603} created their own data set, RLD09, to use in their experiments. To create the data set, the authors captured network data from the Computer Engineering Department at King Mongkut's University of Technology Thonburi, in Bangkok, Thailand. The data has around ten million preprocessed data \emph{packets}. A packet is a unit of data that is carried throughout a network~\cite{networkPacket}. RLD09 has 17 different types of attacks that can be divided into denial of service attacks and probe attacks. It also has normal network activity. RLD09 uses 12 features, which include the number of packets, source ports, and destination ports.

@@

\subsection{Fuzzy Logic}
\label{sec:fuzzyLogic}
Attacks on systems do not always have a fixed pattern, so fuzzy logic is used to detect patterns that have a behavior that is between normal and unusual. A fuzzy logic rule has the following format: If <\emph{condition}> then <\emph{consequence}>, where \emph{condition} is a fuzzy variable, and \emph{consequence} is a fuzzy set. For example: if the number of packets with the same destination address is high, then the pattern is unusual. To determine what is considered high, the values of the packets are divided into fuzzy sets. In Figure~\ref{fig:triangleFigure} there are three sets: low (region A), medium (region B), and high (region C). 
\begin{figure}
\centering
\psfig{file=triangleFigure.JPG}
\caption{Fuzzy shape with three sets: low (A), medium (B), and high (C)~\cite{DBLP:journals/corr/abs-1204-6416}}
\label{fig:triangleFigure}
\end{figure}
The x-axis are the values in the fuzzy set (number of packets), and the y-axis is the membership function, which determines the region. For example, if the number of packets with the same destination address is 15 then this will be considered as low for a degree of 0.4, but will be considered medium or high for other degrees. If the number of packets is determined to be high, then the connection will be aborted.~\cite{DBLP:journals/corr/abs-1204-6416}

@@

\subsection{Algorithm Overview}
In research conducted by M. S. Hoque, M. A. Mukit, and M. A. N. Bikas~\cite{DBLP:journals/corr/abs-1204-1336}, a genetic algorithm was used to make their intrusion detection system. The system is divided into two phases: a precalculation phase and a detection phase. In the precalculation phase, a set of chromosomes are created using training data. Then this set of chromosomes is used in the detection phase for comparison. Algorithm~\ref{alg:precalc} is used in the precalculation phase.

@@

    chromosome}
  \ENDFOR
  \STATE{Remove some chromosomes with worse fitness}
  \STATE{Apply crossover to the selected pair of chromosomes of the population}
  \STATE{Apply mutation to each chromosome of the

  population~//~mutation rate~=~0.35}
\ENDWHILE
\end{algorithmic}
\end{algorithm}

@@

  \STATE{compare the predicted result with actual result}
  \STATE{find $A$, $B$, $\alpha$, and $\beta$}
  \STATE{// $A$ is total number of attack records. $B$ is total number of normal records. $\alpha$ is total number of attack records correctly identified as attack. $\beta$ is total number of normal records incorrectly classified as attack.}
\ENDFOR
\STATE{calculate fitness}
\STATE{//create next generation}
\STATE{preserve\_best()}
\STATE{crossover()}
\STATE{mutation() // mutation rate~=~0.30}
\end{algorithmic}
\end{algorithm}

The fitness function to be maximized is:
\begin{equation*}
\textrm{fitness function} = \frac{\alpha}{A} - \frac{\beta}{B}
\end{equation*}
$A$ is total number of attack records. $B$ is total number of normal records. $\alpha$ is total number of attack records correctly identified as attack. $\beta$ is total number of normal records incorrectly classified as attack.

In this implementation, a population size of 10 was used for each generation. An individual in the population represents a possible detection rule. The two best individuals from a present generation are preserved for the next generation. The other individuals in the new generation come from mutation and single-point crossover.

@@

The authors of~\cite{6496342, 6559603} first trained and tested the fuzzy genetic algorithm (Algorithm~\ref{alg:fuzGenAlg}) with the KDD99 data set. There were 6 different types of denial of service attacks and 4 different types of probe attacks. The detection rate of the KDD99 data set was 98.72\%. Then 26,500 records of the RLD09 data set were used as the training set. The detection rate was 97.97\%. The results of this experiment are shown in Table~\ref{tab:bothSetsResults}.
